Teaching modeling in introductory statistics: A comparison of formula and tidyverse syntaxes

### 2024-07-08 PaperDraft.Rmd
add DOI
another round of revisions

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of pre- and post-survey data show minimal differences between the two labs, with students reporting a positive experience regardless of section. Analysis of data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed that tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials and minutes of the associated YouTube videos. Both labs relied on a relatively small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. The tidyverse labs exposed students to more distinct R functions, but reused functions more frequently. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
biblio-style: apalike
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

@@

```{r blinding, echo=FALSE}
blind <- FALSE
```

@@

our_table <- function(x, ...) {
  kableExtra::kbl(x, ...) %>%
    kableExtra::kable_styling(latex_options = "striped")
}

@@

```{r, cache = FALSE}
library(here)
library(tidyverse)
library(kableExtra)
library(lme4)
library(broom.mixed)
library(bookdown)
library(ggh4x)
library(cowplot)
ggplot2::theme_set(ggplot2::theme_gray(base_size = 14))
```

@@

rstudio_cloud <- read_csv(here("data", "processed", "rstudio_cloud.csv"), col_types = cols(
  month = col_character(),
  amount = col_double(),
  section = col_character()
)) %>%
  mutate(section = if_else(section == "tidy", "tidyverse", section)) %>%
  mutate(month = as_factor(month)) %>%
  mutate(month = fct_relevel(month, "September", "October", "November", "December"))

@@

qs <- prepostsurvey %>%
  slice(1)

prepostsurvey <- prepostsurvey %>%
  slice(-1) %>%
  mutate(
    Section.x = case_when(
      Section.x == "202040STAT220-01 and 202040STAT220-52" ~ "formula",
      Section.x == "202040STAT220-01 and 202040STAT220-53" ~ "tidyverse"
    ),
    Section.y = case_when(
      Section.y == "202040STAT220-01 and 202040STAT220-52" ~ "formula",
      Section.y == "202040STAT220-01 and 202040STAT220-53" ~ "tidyverse"
    )
  )

allfunctions <- read_csv(here("data", "processed", "allfunctions.csv"),
  col_types = cols(
    section = col_character(),
    text = col_character(),
    n = col_double()
  )
) %>%
  mutate(section = if_else(section == "tidy", "tidyverse", section))

csfunctions <- read_csv(here("data", "processed", "csfunctions.csv"),
  col_types = cols(
    section = col_character(),
    text = col_character(),
    n = col_double()
  )
) %>%
  mutate(section = if_else(section == "tidy", "tidyverse", section))

lablines <- read_csv(here("data", "processed", "lablines.csv"),
  col_types = cols(
    file = col_character(),
    lines = col_double(),
    section = col_character()
  )
) %>%
  mutate(section = if_else(section == "tidy", "tidyverse", section))

@@

Many tools for learning were inspired by Rolf Biehler's 1997 paper, "Software for Learning and for Doing Statistics" [@biehler1997]. In it, Biehler called for more attention to the design of tools used for teaching. In particular, he was concerned with on-ramps for students (ensuring the tool was not too complex), as well as off-ramps (using one tool through an entire class, which could also extend further) [@biehler1997]. At the time he wrote the paper it was quite difficult to teach using an authentic tool for doing, because these tools lacked technological or pedagogical on-ramps.

However, recent developments in Integrated Development Environments (IDEs) and pedagogical advances have opened space for a movement to teach even novices statistics and data science using programming. In particular, curricula using Python and R have become popular. In these curricula, educators make pedagogical decisions about what code to show students, and how to scaffold it. In both the Python and R communities, there have been movements to simplify syntax for students.

For example, the UC Berkeley Data 8 course uses Python, including elements of the commonly-used `matplotlib` and `numpy` libraries as well as a specialized library written to accompany the curriculum called `datascience` [@adhikarietal2021; @deneroetal2022]. The `datascience` library was designed to reduce complexity in the code. At the K-12 level, the language `Pyret` has been developed as a simplified version of Python to accompany the Bootstrap Data Science curriculum [@krishnamurthietal2020].

In R, the development of less-complex code for students has been under consideration for even longer [@pruimetal2011]. R offers non-standard evaluation, which allows package authors to create new 'syntax' for their packages [@morandatetal2012]. In human language, syntax is the set of rules for how words and sentences should be structured. If you use the wrong syntax in human language, people will hear there is something wrong with your speech or writing. However, because human understanding is flexible, the listener will probably still understand the general idea you were trying to convey. Syntax in programming language is more formal-- it governs what code will execute, run, or compile correctly. Using the wrong syntax usually means failing to get a result from the program.

@@

The base syntax is used by the base R language [@rcoreteam2020], and is characterized by the use of dollar signs and square brackets. The formula syntax uses the tilde to separate response and explanatory variable(s) [@pruimetal2017]. The `tidyverse` syntax uses a data-first approach, and the pipe to move data between steps [@wickhametal2019].

@@

```{r, eval = FALSE}
library(palmerpenguins)
data("penguins")
# penguins <- penguins %>%
#   drop_na(bill_length_mm)
library(mosaic)
library(tidyverse)
```

```{r r-syntax, ref = "r-syntax", codecap = "Making a histogram of bill length from the penguins dataset, then taking the mean, using three different R syntaxes. Base syntax is characterized by the dollar sign, formula by the tilde, and tidyvese is dataframe-first.", echo = TRUE, eval = FALSE}
# base syntax
hist(penguins$bill_length_mm)
mean(penguins$bill_length_mm, na.rm = TRUE)

# formula syntax
gf_histogram(~bill_length_mm, data = penguins)
mean(~bill_length_mm, data = penguins, na.rm = TRUE)

# tidyverse syntax
ggplot(penguins) +
  geom_histogram(aes(x = bill_length_mm))
penguins %>%
  drop_na(bill_length_mm) %>%
  summarize(mean(bill_length_mm))
```

@@

While there are strongly-held opinions on which syntax should be taught [@pruimetal2017; @cetinkaya-rundeletal2022], there is relatively little empirical evidence to support these opinions. In the realm of computer science, empirical studies by Andreas Stefik, et al have shown significant differences in the intuitiveness of languages, as well as error rates, based on language design choices [@stefiketal2011; @stefiksiebert2013]. Thus, it seems likely there are language choices that could make data science programming easier (or harder) for users, particularly novices.

Rafalski et al (a group which includes Stefik) ran an experiment comparing the three main R syntaxes. The study showed no statistically significant difference between the three syntaxes with regard to time to completion or number of errors. However, there were significant interaction effects between syntax and task, which suggested some syntaxes might be more appropriate for certain tasks [@rafalskietal2019].

Examining the results from the study with an eye toward data science pedagogy showed common errors made by students related to their conceptions of dataframes and variables. For example, one of the figures from @rafalskietal2019 shows real student code with errors. In the first line of code, the student gets everything correct using formula syntax, with the exception of the name of the dataframe. When that code does not work, they try again using base R syntax, but again get the dataframe name wrong. After both those failures, they appear to fall back on computer science knowledge and try syntax quite different from R. This is consistent with other studies of novice behavior in R [@roberts2015]. It is not clear if this type of error was dependent on the syntax participants were asked to use.

The other missing element in this study was instruction. The study was a quick intervention showing students examples of a particular syntax, then asking them to duplicate that syntax in a new situation. But without any instruction about data science concepts like dataframes, it would be difficult to troubleshoot the syntax error mentioned above. The work served as the inspiration for the longer comparison of multiple R syntaxes in the classroom context described in this paper.

In this paper, we seek to more deeply compare the experience of teaching and learning R in introductory statistics classes using the formula and `tidyverse` syntaxes. If you are looking for a definitive answer on which syntax is 'best,' you will be disappointed with the conclusions of this paper. But, if you are an instructor trying to weigh the benefits and drawbacks of each syntax, this paper will provide additional questions and evidence you can use to guide your decision.

@@

Data was collected from YouTube analytics for watch times, from RStudio Cloud for aggregated compute time, and from pre- and post-surveys of students. Participants for the pre- and post-survey were recruited from this pool after Institutional Research Board ethics review (University of St Thomas IRB 1605810-2).

<!-- REDACTED IRB number -->

@@

Participants were students enrolled in two introductory statistics lab sections associated with a common introductory statistics course. The two labs were of the same size ($n=21$ in both sections) and reasonably similar in terms of student composition. In both sections, approximately half of students were Business majors, with the other half a mix of other majors.

@@

```{r, eval=FALSE}
prepostsurvey %>%
  count(Section.x)

prepostsurvey %>%
  drop_na(Q11.y) %>%
  count(Section.y)
```

```{r priorexptab, results='asis'}
table(prepostsurvey$Q1, prepostsurvey$Section.x) %>%
  kable(caption = "Responses from pre-survey about prior programming experience. The majority of students in both sections had no prior programming experience.", booktabs = TRUE)
```

The course these labs were associated with is an introductory statistics course, so it is a first introduction to statistics for all but a handful of students. Typically, students also have no prior experience programming. To verify this, we compared the prior programming experience of the two groups of students. Table \ref{tab:priorexptab} shows results from the pre-survey. While two additional students in the `tidyverse` section had prior programming experience, the majority of students in both sections had no prior programming experience.

@@

The course these labs accompanied was an introductory statistics course using the Lock5 textbook [@locketal2020]. Weekly lab topics were aligned with lecture content, and standardized across all labs for the semester (13 lab sections using a variety of different software). The topics covered in Fall 2020 are shown in Table \ref{topics}.

```{=tex}
\begin{table}
\singlespacing
\begin{tabular}{p{\linewidth}}
1. [No lab, short week] \\ 
2. Describing data: determining the number of observations and variables in a dataset, variable types. \\
3. Categorical variables: exploratory data analysis for one or two categorical variables. Frequency tables, relative frequency tables, bar charts, two-way tables, and side-by-side bar charts. \\
4. Quantitative variables: exploratory data analysis for one quantitative variable. Histograms, dot plots, density plots, and summary statistics like mean, median, and standard deviation. \\
5. Correlation and regression: exploratory data analysis for two quantitative variables. Correlation, scatterplot, simple linear regression as a descriptive technique. \\
6. Bootstrap intervals: the use of the bootstrap to construct non-parametric confidence intervals. \\
7. Randomization tests: the use of randomization to perform non-parametric hypothesis tests. \\
8. Inference for a single proportion: use of the normal distribution to construct confidence intervals and perform hypothesis tests for a single proportion. \\
9. Inference for a single mean: use of the t-distribution to construct confidence intervals and perform hypothesis tests for a single mean. \\
10. Inference for two samples: use of distributional approximations (normal or t) to perform inference for a difference of proportions or a difference of means.\\
11. [No lab, assessment]\\
12. [No lab, Thanksgiving]\\
13. ANOVA: inference for more than two means, using the F distribution.\\
14. Chi-square: inference for more than two counts, using the $\chi^2$ distribution.\\
15. Inference for Regression: inference for the slope coefficient in simple linear regression, prediction and confidence intervals. Multiple regression. \\
Finals. [Second assessment] \\
\end{tabular}
\caption{List of topics for labs across the 15-week semester.}\label{topics}
\end{table}
\doublespacing
```
Although this was a 15-week semester, there are only 12 lab topics. Labs were not held during the first week of classes or during Thanksgiving week. Additionally, there were two "lab assessments" to gauge student understanding of concepts within the context of their lab software. One took place during finals week, the other was scheduled in week 11.

@@

The instructor prepared weekly RMarkdown documents for students, as described in Section \ref{sec:structure}. One question is whether the materials presented to students were of approximately the same length. We can first assess this using the length of the pre-lab RMarkdown documents, measured using lines. Figure \ref{fig:prelablength} shows the number of lines of code and text in each section's pre-lab document, per week. Lines in the RMarkdown document include the YAML header (consistent between documents), the descriptive text about processes (largely similar between documents), and the code in code chunks, which varied based on the syntax of the lab.

```{r prelablength, fig.cap = "Length of pre-lab RMarkdown documents each week, in lines. Data has been adjusted for the formula section in weeks 8 and 9, because an instructor error led this section to have only one document combining both weeks' work."}
lablines <- lablines %>%
  left_join(topics, by = c("file" = "topic")) %>%
  complete(week = full_seq(1:15, 1), nesting(section))

lablines %>%
  ggplot(aes(x = week, y = lines, fill = section)) +
  geom_col(position = "dodge") +
  scale_x_continuous(breaks = seq(from = 1, to = 15, by = 2)) +
  ylab("Length (in lines) of pre-lab \n RMarkdown documents") +
  xlab("Week of semester") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.title.y = element_text(size = 14)
  )
```

Every attempt was made to align these RMarkdown documents, so the descriptive text was only changed when necessary to describe specific elements of the code. Similarly, if blank code chunks appeared in one lab, that was mirrored by a blank chunk in the other lab. Both labs' documents were styled using the `styler` package [@mullerwalthert2022] to remove inconsistencies with spacing, assignment operators, and the like. The `styler` package is based on the tidyverse style guide [@wickham2022], and has become the de facto style guide for R. Previously-existing style guides like the Google R Style Guide have largely decided to follow @wickham2022.

Figure \ref{fig:prelablength} indicates RMarkdown documents for the `tidyverse` section tended to be longer. A slightly longer length for `tidyverse` materials makes sense, because `tidyverse` code is characterized by multiple short lines strung together into a pipeline with `%>%`, while the formula syntax typically has single function calls, sometimes with more arguments. The code shown in this paper is styled the same way as the labs, so any code comparison you see (for example, in \ref{r-syntax} or \ref{tally-ex1} and \ref{tidy-tally1}) will show the difference in lines of code between the two syntaxes.

```{r}
labsfortesting <- lablines %>%
  drop_na(file) %>%
  pivot_wider(week, names_from = section, values_from = lines) %>%
  mutate(
    diff = tidyverse - formula,
    perc_diff = (diff / formula) * 100
  ) %>%
  drop_na(week)
```

@@

```{r videolength, fig.cap = "Length of pre-lab videos each week. Outlines delineate multiple videos for a single week."}
youtube_videos <- youtube_videos %>%
  mutate(week_topic = as_factor(week_topic)) %>%
  mutate(week_topic, fct_reorder(week_topic, week)) %>%
  complete(week = full_seq(1:15, 1), nesting(section))

youtube_videos %>%
  mutate(week = case_when(
    section == "formula" ~ week - 0.15,
    section == "tidyverse" ~ week + 0.25
  )) %>%
  ggplot(aes(x = week, y = tot_min, fill = section)) +
  geom_col(position = "stack", color = "white", width = 0.4) +
  scale_x_continuous(breaks = seq(from = , to = 15, by = 2)) +
  # facet_wrap(~section, nrow = 2) +
  ylab("Total length of pre-lab \n videos (minutes)") +
  xlab("Week of semester") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.title.y = element_text(size = 14)
  )
```

@@

Figure \ref{fig:videolength} shows the lengths of videos over the course of the semester. Video length appears generally consistent between sections. Effort was made to ensure the maximum video length was approximately 20 minutes, although some weeks had multiple videos.

```{r}
youtube_fortesting <- youtube_weeks %>%
  select(week, section, tot_min) %>%
  pivot_wider(week, names_from = section, values_from = tot_min) %>%
  mutate(
    diff = tidyverse - formula,
    avglength = (tidyverse + formula) / 2,
    perc_diff = (diff / formula) * 100
  )
```

We can again compute the percentage difference in total video length (adding together multiple videos in weeks that had them), and compute the mean of that difference. This shows us that `tidyverse` labs were `r round(mean(youtube_fortesting$diff))` minutes longer or `r round(mean(youtube_fortesting$perc_diff))`% longer than formula videos, on average. The `tidyverse` labs appear to be slightly longer in of lines of code and video length, although these differences are slight.

@@

The results from this work can be split into quantitative and qualitative evidence. Quantitative results can be derived from the pre- and post-survey in the class (the most intentionally-collected data), as well as YouTube Analytics and RStudio Cloud usage data (incidental data). We can also quantify the number of functions used in both sections by analyzing the RMarkdown documents as data in their own right. All of these quantitative results show slight differences between sections, which hint toward some of the qualitative differences in experience. We also discuss these qualitative differences by highlighting three topics where the experience teaching and learning the two syntaxes diverged.

@@

````{=tex}
\begin{itemize}
\itemsep-3mm
```{r, results = 'asis'}
qs %>%
  select(Q11.x:Q16.x) %>%
  pivot_longer(cols = everything(), names_to = "old", values_to = "q") %>%
  select(-old) %>%
  mutate(q = paste("\\item", q, "\n")) %>%
  pull(q) %>%
  walk(cat)
```
\end{itemize}
````

>>>>

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of pre- and post-survey data show minimal differences between the two labs, with students reporting a positive experience regardless of section. Analysis of data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials and minutes of the associated YouTube videos. The tidyverse labs exposed students to more distinct R functions, but reused functions more frequently. Both labs relied on a relatively small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
biblio-style: agsm
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

@@

```{r blinding, echo=FALSE}
blind <- TRUE
```

@@

our_table <- function(x, ...) {
  kableExtra::kbl(x, ...) |>
    kableExtra::kable_styling(latex_options = "striped")
}

@@

```{r, cache = FALSE}
library(here)
library(tidyverse)
library(kableExtra)
library(lme4)
library(broom.mixed)
library(bookdown)
library(ggh4x)
library(cowplot)
library(extrafont)
library(readxl)
# font_install('fontcm')
# https://github.com/wch/fontcm
ggplot2::theme_set(ggplot2::theme_gray(base_size = 14))
```

@@

rstudio_cloud <- read_csv(here("data", "processed", "rstudio_cloud.csv"), col_types = cols(
  month = col_character(),
  amount = col_double(),
  section = col_character()
)) |>
  mutate(section = if_else(section == "tidy", "tidyverse", section)) |>
  mutate(month = as_factor(month)) |>
  mutate(month = fct_relevel(month, "September", "October", "November", "December"))

@@

qs <- prepostsurvey |>
  slice(1)

prepostsurvey <- prepostsurvey |>
  slice(-1) |>
  mutate(
    Section.x = case_when(
      Section.x == "202040STAT220-01 and 202040STAT220-52" ~ "formula",
      Section.x == "202040STAT220-01 and 202040STAT220-53" ~ "tidyverse"
    ),
    Section.y = case_when(
      Section.y == "202040STAT220-01 and 202040STAT220-52" ~ "formula",
      Section.y == "202040STAT220-01 and 202040STAT220-53" ~ "tidyverse"
    )
  )

allfunctions <- read_excel("data/processed/allfunctions_packages.xlsx") |>
  mutate(section = if_else(section == "tidy", "tidyverse", section)) |>
  mutate(text = paste0(text, "()"))
# allfunctions <- read_csv(here("data", "processed", "allfunctions.csv"),
#   col_types = cols(
#     section = col_character(),
#     text = col_character(),
#     n = col_double()
#   )
# # ) |>
#   mutate(section = if_else(section == "tidy", "tidyverse", section))

csfunctions <- read_csv(here("data", "processed", "csfunctions.csv"),
  col_types = cols(
    section = col_character(),
    text = col_character(),
    n = col_double()
  )
) |>
  mutate(section = if_else(section == "tidy", "tidyverse", section))

lablines <- read_csv(here("data", "processed", "lablines.csv"),
  col_types = cols(
    file = col_character(),
    lines = col_double(),
    section = col_character()
  )
) |>
  mutate(section = if_else(section == "tidy", "tidyverse", section))

@@

Many tools for learning were inspired by Rolf Biehler's 1997 paper, "Software for Learning and for Doing Statistics." In it, Biehler called for more attention to the design of tools used for teaching. In particular, he was concerned with on-ramps for students (ensuring the tool was not too complex), as well as off-ramps (using one tool through an entire class, which could also extend further) [@biehler1997]. At the time he wrote the paper it was quite difficult to teach using an authentic tool for doing, because these tools lacked technological or pedagogical on-ramps.

However, recent Integrated Development Environments (IDEs) and pedagogical advances have opened space for a movement to teach even novices statistics and data science using programming. In particular, curricula using Python and R have become popular. In these curricula, educators make pedagogical decisions about what code to show students, and how to scaffold it. In both the Python and R communities, there have been movements to simplify syntax for students.

For example, the UC Berkeley Data 8 course uses Python, including elements of the commonly-used `matplotlib` and `numpy` libraries as well as a specialized library written to accompany the curriculum called `datascience` [@adhikarietal2021; @deneroetal2022]. The `datascience` library was designed to reduce complexity in the code. At the K-12 level, the language Pyret has been developed as a simplified version of Python to accompany the Bootstrap Data Science curriculum [@krishnamurthietal2020].

In R, the development of less-complex code for students has been under consideration for even longer [@pruimetal2011]. R offers non-standard evaluation, which allows package authors to create new 'syntax' for their packages [@morandatetal2012]. In human language, syntax is the set of rules for how words and sentences should be structured. If someone uses the wrong syntax in human language, people will hear there is something wrong with the speech or writing. However, because human understanding is flexible, the listener will probably still understand the general idea the speaker was trying to convey. Syntax in programming language is more formal-- it governs what code will execute, run, or compile correctly. Using the wrong syntax usually means failing to get a result from the program.

@@

The base syntax is used by the base R language [@rcoreteam2020], and is characterized by the use of dollar signs and square brackets. The formula syntax uses the tilde to separate response and explanatory variable(s) [@pruimetal2017]. The `tidyverse` syntax uses a data-first approach, and the pipe to move data between steps [@wickhametal2019]. In this paper, the base R pipe `|>` has been used, but in the original pedagogical materials students saw the `magrittr` pipe, `%>` [@bachewickham2014].

@@

```{r r-syntax, ref = "r-syntax", codecap = "Making a histogram of bill length from the penguins dataset, then taking the mean, using three different R syntaxes. Base syntax is characterized by the dollar sign, formula by the tilde, and tidyvese is dataframe-first.", echo = TRUE, eval = FALSE}
# load data
library(palmerpenguins)
data("penguins")

# base syntax
hist(penguins$bill_length_mm)
mean(penguins$bill_length_mm, na.rm = TRUE)

# formula syntax
library(mosaic)
gf_histogram(~bill_length_mm, data = penguins)
mean(~bill_length_mm, data = penguins, na.rm = TRUE)

# tidyverse syntax
library(tidyverse)
ggplot(penguins) +
  geom_histogram(aes(x = bill_length_mm))
penguins |>
  drop_na(bill_length_mm) |>
  summarize(mean(bill_length_mm))
```

@@

While there are strongly-held opinions on which syntax should be taught [@pruimetal2017; @cetinkaya-rundeletal2022], there is relatively little empirical evidence to support these opinions. In the realm of computer science, empirical studies have shown significant differences in the intuitiveness of languages, as well as error rates, based on language design choices [@stefiketal2011; @stefiksiebert2013]. Thus, it seems likely there are language choices that could make data science programming easier (or harder) for users, particularly novices.

Rafalski et al ran an experiment comparing the three main R syntaxes. The study showed no statistically significant difference between the three syntaxes with regard to time to completion or number of errors. However, there were significant interaction effects between syntax and task, which suggested some syntaxes might be more appropriate for certain tasks [@rafalskietal2019].

Examining the results of the study with an eye toward data science pedagogy showed common errors related to conceptions of dataframes and variables. For example, one of the figures from @rafalskietal2019 shows real student code with errors. In the first line of code, the student gets everything correct using formula syntax, with the exception of the name of the dataframe. When that code does not work, they try again using base R syntax, but again get the dataframe name wrong. After both those failures, they appear to fall back on computer science knowledge and try syntax quite different from R. This is consistent with other studies of novice behavior in R [@roberts2015]. It is not clear if this type of error was dependent on the syntax participants were asked to use.

The study was a quick intervention showing students examples of a particular syntax, then asking them to duplicate that syntax in a new situation. But without any instruction about data science concepts like dataframes, it would be difficult to troubleshoot the syntax error mentioned above. The work served as the inspiration for the longer comparison of multiple R syntaxes in the classroom context described in this paper.

In this paper, we seek to more deeply compare the experience of teaching and learning R in an introductory statistics class using the formula and `tidyverse` syntaxes. This manuscript does not provide an answer to which syntax is 'best,' but instead helps to describe the benefits and drawbacks of each syntax. This paper provides instructors with additional questions to ask themselves and evidence to use in service of making a decision on syntax.

@@

Data was collected from YouTube analytics for watch times, from RStudio Cloud for compute time, and from pre- and post-surveys of students. Participants for the pre- and post-survey were recruited from this pool after Institutional Research Board ethics review (University of St Thomas IRB 1605810-2).

<!-- (REDACTED IRB number).  -->

@@

Participants were students enrolled in two introductory statistics lab sections associated with a common introductory statistics course. The two labs were of the same size ($n=21$ in both sections) and reasonably similar in terms of student composition. In both sections, approximately half of students were business majors, with the other half a mix of other majors.

@@

```{r, eval=FALSE}
prepostsurvey |>
  count(Section.x)

prepostsurvey |>
  drop_na(Q11.y) |>
  count(Section.y)
```

```{r priorexptab, results='asis'}
priorexp <-  prepostsurvey |>
  count(Q1, Section.x)
# table(prepostsurvey$Q1, prepostsurvey$Section.x) |>
#   kable(caption = "Responses from pre-survey about prior programming experience. The majority of students in both sections had no prior programming experience.", booktabs = TRUE)
```

This lab course accompanied an introductory statistics course, a first introduction to statistics for all but a handful of students. Typically, students also have no prior experience programming. In the formula section, `r filter(priorexp, Q1=="No", Section.x == "formula") |> pull(n)` had no prior experience programming, and `r filter(priorexp, Q1=="Yes, but not with R", Section.x == "formula") |> pull(n)` had experience programming, but not with R. In the `tidyverse` section, `r filter(priorexp, Q1=="No", Section.x == "tidyverse") |> pull(n)` had no prior experience programming, and `r filter(priorexp, Q1=="Yes, but not with R", Section.x == "tidyverse") |> pull(n)` had experience programming, but not with R. While two additional students in the `tidyverse` section had prior programming experience, the majority of students in both sections had no prior programming experience.

@@

The course these labs accompanied was an introductory statistics course using the Lock5 textbook [@locketal2020]. Weekly lab topics were aligned with lecture content, and standardized across all labs for the semester (13 lab sections using a variety of different software). The topics covered are shown in Table \ref{topics}.

```{=tex}
\begin{table}
\singlespacing
{\setlength{\extrarowheight}{0.1cm}
\begin{tabular}{p{1cm}|p{15cm}}
Week & Topic \\
1 & \em{No lab, short week} \\  
2 & Describing data: determining the number of observations and variables in a dataset, variable types. \\  
3 & Categorical variables: exploratory data analysis for one or two categorical variables. Frequency tables, relative frequency tables, bar charts, two-way tables, and side-by-side bar charts. \\  
4 & Quantitative variables: exploratory data analysis for one quantitative variable. \newline Histograms, dot plots, density plots, and summary statistics like mean, median, and standard deviation. \\  
5 & Correlation and regression: exploratory data analysis for two quantitative variables. Correlation, scatterplot, simple linear regression as a descriptive technique. \\  
6 & Bootstrap intervals: the use of the bootstrap to construct non-parametric confidence intervals. \\  
7 & Randomization tests: the use of randomization to perform non-parametric \linebreak hypothesis tests. \\  
8 & Inference for a single proportion: use of the normal distribution to construct 
\linebreak confidence intervals and perform hypothesis tests for a single proportion. \\  
9 & Inference for a single mean: use of the t-distribution to construct confidence intervals and perform hypothesis tests for a single mean. \\  
10 & Inference for two samples: use of distributional approximations (normal or t) to perform inference for a difference of proportions or a difference of means.\\  
11 & \em{No lab, assessment}\\  
12 & \em{No lab, Thanksgiving}\\  
13 & ANOVA: inference for more than two means, using the F distribution.\\  
14 & Chi-square: inference for more than two counts, using the $\chi^2$ distribution.\\  
15 & Inference for regression: inference for the slope coefficient in simple linear regression, prediction and confidence intervals. Multiple regression. \\  
Finals & \em{Second assessment} \\  
\end{tabular}
}
\caption{List of topics for labs across the 15-week semester.}\label{topics}
\end{table}
\doublespacing
```
Although this was a 15-week semester, there are only 12 lab topics. Labs were not held during the first week of classes or during Thanksgiving week. Additionally, there were two "lab assessments" to gauge student understanding of concepts within the context of their lab software. One took place during finals week, the other was scheduled in week 11.

@@

The instructor prepared weekly RMarkdown documents for students, as described in Section \ref{sec:structure}. One question is whether the materials presented to students were of approximately the same length. We can first assess this using the length of the pre-lab RMarkdown documents, measured using lines. Figure \ref{fig:prelablength}a shows the number of lines of code and text in each section's pre-lab document, per week. Lines in the RMarkdown document include the YAML header (consistent between documents), the descriptive text about processes (largely similar between documents), and the code in code chunks, which varied based on the syntax of the lab.

```{r prelablength, fig.cap="Length of pre-lab documents, in lines and minutes.", fig.subcap = c("Length of pre-lab RMarkdown documents each week, in lines. Data has been adjusted for the formula section in weeks 8 and 9, because an instructor error led this section to have only one document combining both weeks' work.", "Length of pre-lab videos each week. Outlines delineate multiple videos for a single week."), fig.ncol = 1, fig.align='left',fig.sep = rep("\\newline", 2)}
lablines <- lablines |>
  left_join(topics, by = c("file" = "topic")) |>
  complete(week = full_seq(1:15, 1), nesting(section))

p1 <- lablines |>
  ggplot(aes(x = week, y = lines, fill = section)) +
  geom_col(position = "dodge") +
  scale_x_continuous(breaks = seq(from = 1, to = 15, by = 2)) +
  ylab("Length (in lines) of pre-lab \n RMarkdown documents") +
  xlab("Week of semester") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.title.y = element_text(size = 14)
  )

youtube_videos <- youtube_videos |>
  mutate(week_topic = as_factor(week_topic)) |>
  mutate(week_topic, fct_reorder(week_topic, week)) |>
  complete(week = full_seq(1:15, 1), nesting(section))

p2 <- youtube_videos |>
  mutate(week = case_when(
    section == "formula" ~ week - 0.15,
    section == "tidyverse" ~ week + 0.25
  )) |>
  ggplot(aes(x = week, y = tot_min, fill = section)) +
  geom_col(position = "stack", color = "white", width = 0.4) +
  scale_x_continuous(breaks = seq(from = , to = 15, by = 2)) +
  # facet_wrap(~section, nrow = 2) +
  ylab("Total length of pre-lab \n videos (minutes)") +
  xlab("Week of semester") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.title.y = element_text(size = 14)
  )
p1
p2

# plot_grid(p1, p2, nrow=2)
```

Every attempt was made to align these RMarkdown documents, so the descriptive text was only changed when necessary to describe specific elements of the code. Similarly, if blank code chunks appeared in one lab, that was mirrored by a blank chunk in the other lab. Both labs' documents were styled using the `styler` package [@mullerwalthert2022] to remove inconsistencies with spacing, assignment operators, and the like. The `styler` package is based on the `tidyverse` style guide [@wickham2022], and has become the de facto style guide for R. Previously-existing style guides like the Google R Style Guide have largely decided to follow @wickham2022.

Figure \ref{fig:prelablength}a indicates RMarkdown documents for the `tidyverse` section tended to be longer. A slightly longer length for `tidyverse` materials makes sense, because `tidyverse` code is characterized by multiple short lines strung together into a pipeline with `|>`, while the formula syntax typically has single function calls, sometimes with more arguments. (When students engaged with this material, they were using the `magrittr` pipe, `%>%`, but code shown here has been adjusted to the new native base R pipe, `|>`. This does not have an impact on the number of lines of code.) The code shown in this paper is styled the same way as the labs, so any code comparison shown (for example, in \ref{r-syntax} or \ref{tally-ex1} and \ref{tidy-tally1}) will show the difference in lines of code between the two syntaxes.

```{r}
labsfortesting <- lablines |>
  drop_na(file) |>
  pivot_wider(week, names_from = section, values_from = lines) |>
  mutate(
    diff = tidyverse - formula,
    perc_diff = (diff / formula) * 100
  ) |>
  drop_na(week)
```

@@

@@

Figure \ref{fig:prelablength}b shows the lengths of videos over the course of the semester. Video length appears generally consistent between sections. Effort was made to ensure the maximum video length was approximately 20 minutes, although some weeks had multiple videos.

```{r}
youtube_fortesting <- youtube_weeks |>
  select(week, section, tot_min) |>
  pivot_wider(week, names_from = section, values_from = tot_min) |>
  mutate(
    diff = tidyverse - formula,
    avglength = (tidyverse + formula) / 2,
    perc_diff = (diff / formula) * 100
  )
```

We can again compute the percentage difference in total video length (adding together multiple videos in weeks that had them), and compute the mean of that difference. This shows us that `tidyverse` labs were `r round(mean(youtube_fortesting$diff))` minutes longer or `r round(mean(youtube_fortesting$perc_diff))`% longer than formula videos, on average. The `tidyverse` labs appear to be slightly longer in terms of lines of code and video length, although these differences are slight.

@@

The results from this work include both quantitative and qualitative evidence. Quantitative results can be derived from the pre- and post-survey in the class (intentionally-collected data), as well as YouTube Analytics and RStudio Cloud usage data (incidental data). We can also quantify the number of functions used in both sections by analyzing the RMarkdown documents as data in their own right. All of these quantitative results show slight differences between sections, which hint toward some of the qualitative differences in experience. We also discuss these qualitative differences by highlighting three topics where the experience teaching and learning the syntaxes diverged.

@@

````{=tex}
\begin{itemize}
\itemsep-3mm
```{r, results = 'asis'}
qs |>
  select(Q11.x:Q16.x) |>
  pivot_longer(cols = everything(), names_to = "old", values_to = "q") |>
  select(-old) |>
  mutate(q = paste("\\item", q, "\n")) |>
  pull(q) |>
  walk(cat)
```
\end{itemize}
````

Figure \ref{fig:pre-post} displays a visualization of these Likert-scale questions.The visualization shows the pre- and post-survey responses to the survey questions, broken down by section. There is no obvious overall trend. However, considerable improvement was seen in the Programming Confident category, which suggests that students in both sections became more confident in their "ability to make use of programming software to work with data."
